---
title: "Class 19: Pertussis Mini Proj"
author: "Nathaniel Lightle (A16669288)"
format: pdf
---

Pertussis is a sever lung infection also known as whooping cough. 

We will begin by investigating the number of Pertussis cases per year in the US. 

This data is available on the CDC website[here](https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html)

# 1. Investigating pertussis cases by year

```{r}
#| echo=FALSE
cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L,2020L,2021L),
  Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617,6124,
                                   2116)
)
```

Lets have a wee look at this data table

```{r}
head(cdc)
```

>[Q1]

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(Year, Cases) +
  geom_point() +
  geom_line()
```

# 2. A tale of two vaccines (wP & aP)

>[Q2]

```{r}
ggplot(cdc) + 
  aes(Year, Cases) + 
  geom_point() + 
  geom_line() +
  geom_vline(xintercept = 1946, linetype = "dashed", col = "blue") + 
  geom_vline(xintercept = 1996, linetype = "dashed", col = "red") +
  geom_vline(xintercept = 2020, linetype = "dashed", col = "darkgreen")
```

>[Q3]

After the introduction of the aP vaccine pertussis cases go up again. This could be due to anti-vaccine sentiments spurred up by the covid pandemic.

# Exploring CMI-PB data

The CMI-PB project makes its data available via "API-endpoint" that return JSON format. 
We will use the 'jsonlite' package to access this data. The main function in this package is called `read.json()`. 

```{r}
library(jsonlite)

# Subject table
subject <- read_json("http://cmi-pb.org/api/subject", simplifyVector = TRUE)
head(subject)
#Specimen table
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = TRUE)
#Titer table
titer <- read_json("http://cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = TRUE)
```

Have a wee peak at these new objects
```{r}
head(titer)
```

>[Q4]

```{r}
table(subject$infancy_vac)
```
There are 60 aP and 58 wP.

>[Q5]

```{r}
table(subject$biological_sex)
```
There are 79 females and 39 males.

>[Q6]

```{r}
table(subject$race, subject$biological_sex)
```

There is 1 native american male, 21 asian females, 11 asian males, 2 black females, 9 more-than-one-race females, 2 more-than-one-race males, 1 pacific islander female, 1 pacific islander male, 11 unreported females, 4 unreported males, 35 white females, and 20 white males. 

# Side-Note: Working w/ Dates

```{r}
library(tidyverse)
```

```{r}
today()-mdy("09-14-2002")
```

```{r}
time_length( today() - ymd("2002-09-14"),  "years")
```

>[Q7]

```{r}
subject$age <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
subject$age_years <- time_length(subject$age, "years")
ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )

wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )

```

>[Q8]

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

>[Q9]

```{r}
ggplot(subject) +
  aes(age_years,
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```


# Merge or join tables

>[Q9]

```{r}
meta <- inner_join(specimen, subject)
head(meta)
```

Antibody measurements in the blood

>[Q10]

```{r}
abdata <- inner_join(titer, meta)
head(abdata)
```

>[Q11] How many isotypes are we measuring for all these individuals?

```{r}
table(abdata$isotype)
```

>[Q12]

```{r}
table(abdata$dataset)
```
The # of rows for the most recent dataset was a lot lower.

Lets focus on one of these: IgG
```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

Box plot of MFI_normalised vs antigen

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot()
```

```{r}
head(igg)
```

Focus in on IgG to the Pertussis Toxin (PT) antigen in the 2021 dataset
```{r}
igg.pt <- igg %>% filter(antigen == "PT", dataset == "2021_dataset")
```

```{r}
ggplot(igg.pt) + 
  aes(planned_day_relative_to_boost, 
      MFI_normalised, 
      col = infancy_vac,
      group = subject_id) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0, linetype = "dashed", col ="black") +
  geom_vline(xintercept = 14, linetype = "dashed", col="black")
```

>[Q19]

I notice that the gene expression is at its maximum level on day 14. I also notice that the maximum is significantly greater for wP individuals. 

