---
title: "Class 16 Unix Quarto"
author: "Nathaniel Lightle (A16669288)"
format: pdf
---

Instance Code (i-0fb15c55e88b7b5fb)

ssh -i "bioinf_nathaniel.pem" ubuntu@ec2-35-86-212-61.us-west-2.compute.amazonaws.com

## Questions

>Q1. What does the star character accomplish here? 

The star character is essentially a variable, anything can replace it.

>Q2. How many sequences are in this mouse.1.protein.faa file?

There are 56,158 proteins in the mouse 1 file. 

>Q3. What happens if you run the above command without the > mm-first.fa part?

It doesn't save it to a file. 

>Q4. What happens if you were to use two ‘>’ symbols (i.e. >> mm-first.fa)?

If you use two it adds to the file instead of overwriting it.

>Q5. How would you determine how many sequences are in the mm-second.fa file?

I could use grep. There are 98 sequences.


# HOMEWORK

Making a plot of the downloaded .tsv file.

Downloading my data...
```{r}
zebra <- read.delim("mm-second.x.zebrafish.tsv")
```

Organizing the data
```{r}
colnames(zebra) <- c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")
head(zebra)
```

Making a histogram
```{r}
hist(zebra$bitscore, breaks = 30)
```

Plotting relationship between pident and bitscore
```{r}
plot(zebra$pident  * (zebra$qend - zebra$qstart), zebra$bitscore)
```

Now w/ ggplot
```{r}
library(ggplot2)
ggplot(zebra, aes(pident, bitscore)) + geom_point(alpha=0.1) 
```

Adding a trend line
```{r}
ggplot(zebra, aes((pident * (qend - qstart)), bitscore)) + geom_point(alpha=0.1) + geom_smooth() +
  labs( 
    x = "Percent Identity x Length",
    y = "BitScore",
    title = "Zebrafish Data"
    )
```


