---
title: "Class09: Halloween"
author: "Nathaniel Lightle (A16669288)"
format: pdf
---

Today is Halloween and we will apply lots of the analysis methods and R graphics appraoches to find out all about typical Halloween candy. 

# Importing candy data

```{r}
candyphile = read.csv("candy-data.csv", row.names = 1)
head(candyphile)
```

>[Q1] How many different candy types are in this data set?

```{r}
nrow(candyphile)
```
There are 85 different types of candy in the data set

>[Q2] How many fruity candy types are in the dataset? 

```{r}
sum(candyphile[,2])
```
There are 38 fruity candy types in the dataset

# What is your favorite candy?

>[Q3] What is your favorite candy in the dataset and what is it's win percent?

My favorite candy is Werther's Original's Caramel
```{r}
candyphile["Werther's Original Caramel", "winpercent"]
```
The win percent for Werther's Original Caramel is  41.90431%

>[Q4] What is the winpercent value for “Kit Kat”?

```{r}
candyphile["Kit Kat", "winpercent"]
```
The win percent for Kit Kat is 76.7686%

>[Q5] What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candyphile["Tootsie Roll Snack Bars", "winpercent"]
```
The win percent for Tootsie Roll Snack Bars is 49.6535%

Trying the `skim()` function

```{r}
#install.packages("skimr")
library("skimr")
skim(candyphile)
```

>[Q6] Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The win percent is on a different scale because it ranges from 0-100%

>[Q7] What do you think a zero and one represent for the candy$chocolate column?

A 0 means false and a 1 means true

>[Q8]  Plot a histogram of winpercent values

```{r}
hist(candyphile[,"winpercent"])
```

>[Q9] Is the distribution of winpercent values symmetrical?

No the distribution is skewed to the left

> [Q10] Is the center of the distribution above or below 50%?

The center is below 50%

> [Q11] On average is chocolate candy higher or lower ranked than fruity candy?

```{r}
mean(candyphile$winpercent[as.logical(candyphile$chocolate) == T])
mean(candyphile$winpercent[as.logical(candyphile$fruity) == T])
```

On average chocolate candy is higher ranked than fruity candy.

> [Q12] Is this difference statistically significant?

```{r}
t.test(candyphile$winpercent[as.logical(candyphile$chocolate) == T], candyphile$winpercent[as.logical(candyphile$fruity) == T])
```
The difference is statistically significant because the p value is teeny tiny

# Overall Candy Rankings

Ordering data by winpercent

```{r}
head(candyphile[order(candyphile$winpercent),], n=5)
```

>[Q13] What are the five least liked candy types in this set?

The 5 least liked candy types in the dataset are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters 

>[Q14] What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candyphile[order(candyphile$winpercent),], n=5)
```

The top 5 candies in the dataset are Snickers, Kit Kat, Twix, Reese's Miniatures, and Reese's Peanut Butter Cup

>[Q15] Make a first barplot of candy ranking based on winpercent values.

```{r}
mycols <- rep("gray", nrow(candyphile))
#mycols[2:5] <- "red"
mycols[as.logical(candyphile$fruity) == T] <- "red"
mycols[as.logical(candyphile$chocolate) == T] <-"chocolate"
mycols[as.logical(candyphile$caramel) == T] <- "yellow"
mycols[as.logical(candyphile$peanutyalmondy) == T] <- "blue"
mycols[as.logical(candyphile$nougat) == T] <- "green"
mycols[as.logical(candyphile$crispedricewafer) == T] <- "black"
mycols
```
```{r}
library(ggplot2)
ggplot(candyphile) +
  aes(winpercent, rownames(candyphile)) +
  geom_col(fill=mycols)
```


>[Q16] This is quite ugly, use the reorder() function to get the bars sorted by winpercent

```{r}
library(ggplot2)
ggplot(candyphile) +
  aes(winpercent, reorder(rownames(candyphile),winpercent)) +
  geom_col(fill=mycols)
```

>[Q17] What is the worst ranked chocolate candy?

Sixlets is the worst ranked chocolate candy

>[Q18] What is the best ranked fruity candy?

Starbusts is the best ranked fruity candy

# Taking a look at pricepercent

Looking at value
Plotting pricepercent vs winpercent

```{r}
library(ggrepel)

ggplot(candyphile) +
  aes(winpercent, pricepercent, label=rownames(candyphile)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, size=3.3, max.overlaps = 5)
```

>[Q19] Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures give you the most bang for your buck

>[Q20] What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
tail(candyphile[order(candyphile$pricepercent),], n=5)
```

The top 5 most expensive candy types in the dataset are Hershey's Special Dark, Mr Good Bar, Ring pop, Nik L nip, and Nestle Smarties.

# Exploring the correlation structure

Installing corrplot

```{r}
#install.packages("corrplot")
```

Using corrplot

```{r}
library(corrplot)
cij <- cor(candyphile)
corrplot(cij)
```
>[Q22] Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate are negatively correlated

>[Q23] Similarly, what two variables are most positively correlated?

Winpercent and chocolate are the most highly correlated

# PCA Analysis

```{r}
pca <- prcomp(candyphile, scale = TRUE)
summary(pca)
```

Plotting it

```{r}
plot(pca$x[,1:2])
```

Giving it some color

```{r}
plot(pca$x[,1:2], col=mycols, pch=16)
```

Making new data set of PCA

```{r}
mydata <- cbind(candyphile, pca$x[,1:3])
```

ggPlotting it

```{r}
p <- ggplot(mydata) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(mydata),
            label=rownames(mydata)) +
        geom_point(col=mycols)

p
```
Making labels

```{r}
library(ggrepel)

#p + geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

Using plotly

```{r}
#install.packages("plotly")
library(plotly)
#ggplotly(p)
```
Correlation check
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>[Q24] What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity and pluribus are the most strongly correlated in the positive direction. These make sense to me. 
